A Reproducible Data Analysis Workflow with RMarkdown, Git, Make & Docker

### 2019-11-11 manuscript.Rmd
added thanks to JD in author notes
add changes from Julia Delius

---
title             : "A Reproducible Data Analysis Workflow with R Markdown, Git, Make, & Docker"
shorttitle        : "Reproducible Data Analysis Workflow"

@@

affiliation:
  - id            : "1"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "2"
    institution   : "Humboldt-Universität zu Berlin"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany & London, U.K."

authornote: This paper is fully reproducible using the workflow described in this paper. All materials can be found on the accompanying GitHub repository <https://github.com/aaronpeikert/reproducible-research/>.  To provide feedback or any other type of comment or questions regarding our workflow, please add an issue to the GitHub repository of our paper [https://github.com/aaronpeikert/reproducible-research/issues](https://github.com/aaronpeikert/reproducible-research/issues).

@@

In this tutorial, we describe a workflow to ensure long-term and cross-platform reproducibility of
R-based data analyses. Reproducibility means to us that statistical results together with their generating computational workflows are reported fully, and transparently, and remain sustainably available, such that a given workflow can be re-run by a different person later in time and the results will be identical to the ones intially reported [@herouxCompatibleReproducibilityTaxonomy2018; @claerboutElectronicDocumentsGive1992]. The need to ensure reproducibility directly follows from commonly accepted rules of good scientific practice (such as the guidelines of the German Research Foundation; @dfg2019). Ensuring reproducibility is a prerequisite for replicability and a means to increase trustworthiness of empirical results [@epskamp2019rep].

Here, we combine four software tools, whose interplay can guarantuee full computational reproducibility of data analyses and their reporting. There are various ideas to enhance reproducibility [@piccoloToolsTechniquesComputational2016], of which we believe four of them are particularly important: literate programming: [@ruleTenSimpleRules2019], version control [@barbaHardRoadReproducibility2016], dependency management [@askrenUsingMakeReproducible2016] and containerization [@clyburne-sherinComputationalReproducibilityContainers2018]. We argue that only a workflow using all four concepts in unison can guarantuee confidence in reproducing a scientific report. Even though there are various implementations of these concepts, we believe that for the R environment [@r2019], the following four implementations are best suited. For literate programming, R Markdown [@rmarkdown2019], for version control, Git [@chaconProGit2014], for dependency management, Make [@feldmanMakeProgramMaintaining1979], and for containerization, Docker [@merkelDockerLightweightLinux2014]. Each of these software solutions serves a meta-scientific valuable goal (reproducibility) and increases the productivity of the researcher. They all are highly flexible and powerful, so to master them fully, a considerable amount of practice is necessary. However, for our purposes, it is sufficient to master a minimal valuable subset of functions to ensure reproducibility of scientific analyses.

@@

This paper is fully reproducible using the workflow described in this paper. All materials, including the manuscript source and all R codes and files used to generate this manuscript, can be found on the accompanying GitHub repository [https://github.com/aaronpeikert/reproducible-research/](https://github.com/aaronpeikert/reproducible-research/).

@@

The first step towards reproducibility is to create an R script or R Markdown file as primary entry point for the analysis, which runs on a local computer without error and performs the main statistical analyses. Next, one has to make sure that all files relevant for the analysis can be moved to another computer. To this end, it is recommendable that all files are within one folder (or, subfolders within it) and all paths are relative to that folder. A robust solution to the problem of making sure that file access does not break across computing platforms are [Rstudio projects](https://r4ds.had.co.nz/workflow-projects.html) and the `here` [@here2017] package.

@@

The folder, where all files reside that you need for analysis (code and data), is referred to as a 'project' (or sometimes 'research compendium'). Working with projects is particularly comfortable with RStudio, an integrated development environment (IDE) for R. It is useful to organize a data-analysis project in way that strictly segregates (raw) data and code by placing them in directories called `data` and `R` [see Section 4 in @marwickPackagingDataAnalytical2018]; there are also tools that automatize the standardized creation of folder structures such as `workflowr` [@workflowr].

Sometimes external requirements make it impossible that the data can be stored and shared together with the scripts. In most of the cases we have seen, these are either space constraints or privacy considerations. In these cases, of course, unrestricted reproducibility is not guaranteed. If dividing data and scripts is unavoidable, we recommend to validate all data files using checksums (also called a "hash", e.g., using the functions provided in package `digest` [@digest2019]) before analyzing them. To do this, a checksum must be created and stored at the time of the original analysis. At the time of reproduction, the current checksum must be compared with the stored checksum to ensure data integrity.

@@

The translation of computational results into a human-readable summary, for example into a technical report, a presentation, or a journal manuscript, is time-consuming and error-prone. Typical errors result from copy\&paste mistakes, erroneous rounding, or missed updates of the manuscript when the computer code has changed. To not only create fully reproducible results but also fully reproducible reports, one needs to resort to the Literate Programming paradigm [@knuth1984literate], in which human-readable language and computer-code are interspersed to create dynamic documents whose order follows the logic of thought rather then the order of the computer. R Markdown is a simple markup language to create dynamic documents with embedded chunks of R code that can be exported to standard formats such as documents (docx, rtf, epub), presentations (ppt, html) or websites (html) using the knitr package [@xie2015; @xie2019]. Several packages extend the functionality of knitr. Of particular note are the `papaja` package [@papaja2018], which offers additional functions to enable APA-conformable document formatting, including a journal-style final typeset format, and the `stargazer` package [@stargazer2018], which provides journal-ready tables and reports of statistical models.  Figure 1 illustrates R Markdown syntax using the papaja package and Figure 2 shows the resulting rendered document.

@@

Fundamentally, reproducibility means that results remain identical if neither script nor data have changed. It is often not trivial to find out whether anything in a project has changed over time and if so to "go back in time." The program Git enables you to do both. A good mental model for Git is that it takes a sequence of snapshots of all files it is supposed to track. In the language of Git, these snapshots are "commits".  A commit represents a complete copy of the state of all tracked files. Each commit has a unique identifier (a hash) and a human-readable description (commit message). Going back to one state is as easy as traversing the history of all commits and switch to a given one; it is possible to visually track changes between different versions. Finding a specific version can be made even easier by tagging snapshots, e.g., as "submission," "preprint," or "published." This collection of snapshots is called a "repository," which ideally tracks your entire R project.

@@

To keep track of all changes on your local computer, you only need to use `git add` & `git commit` or `git commit -a` to do both. These commands need to be executed in the terminal, which you can access from within Rstudio (`Shift + Alt + R`).  Rstudio also offers a graphical user interface for Git and, quite likely, this interface is sufficient for almost every interaction with Git.

@@

Git also makes it particularly easy to share and collaborate on a project with other researchers. A popular service for sharing materials via Git is [GitHub](https://github.com). Just sharing Git repositories with the public is always free, private repositories (only visible to persons you invite) are [free for researchers](https://help.github.com/en/articles/applying-for-an-educator-or-researcher-discount). After creating a user account, one can create a new repository where GitHub advises you how to upload your repository from the terminal, e.g., for our repository (here with user name 'aaronpeikert' and repository name 'reproducible-research'):

@@

`git push` or the green up-arrow in the git pane (see Figure 3) uploads local updates. To download this git repo to another computer type into the terminal:

@@

Git and GitHub can do even more to support you collaborating with fellow researchers, for example, by providing a web interface to track issues and their status (open/closed/resolved) and further means to managing and merging multiple, parallel versions of code (such as branches, pull requests, or merges) but this is beyond the scope of this tutorial. In particular, GitHub's issue management can be leveraged as post-publication platform to discuss manuscripts and their results [^1].
Another benefit of using Git and GitHub is that experimentation is highly encouraged since you can go back to any state quickly and even when you lose access to the file on your computer, everything is backed up on the GitHub servers. Further, one can reduce the chance of accumulating dead code (e.g., commented out lines) because it is safe to simply remove unneeded code blocks and track the removal in git.

@@

Even when you have obtained a given version of a project with the aim to reproduce reported result and you can ensure that this version is unchanged, you may not know exactly how to reproduce the results, because it may be unclear which files should be executed in which order. This is particularly the case when complex preprocessing pipelines are part of the computation and/or external dependencies are many. Handling such dependencies is comfortable with Make because it allows you to create (computational) recipes to (re-)create files.

@@

The first line can be interpreted as follow: To create `analysis.pdf`, the file `analysis.Rmd` needs to be rendered, which depends on `data/clean.csv`. This dependency is itself a target. To create `data/clean.csv`, `R/clean.R` and `data/iris.csv` are needed. If you type `make analysis.pdf`, `make` first checks whether the dependencies do exist and, if not, creates them. Here, if `data/clean.csv` does not exist, `make` creates it. Also, if one of the dependencies of a target is newer then the target itself, then Make updates everything that directly or indirectly depends on the target. Here, if `data\iris.csv` is newer then `data\clean.csv`, Make will attempt to recreate `data\clean.csv` first and then `analysis.pdf`. If there is a dependency which is missing, and there is no recipe to make it, `make` stops with an error. It is a convention to have a target `all` which depends on everything and make it the first target in a `Makefile`, then the command `make` without any argument automatically creates everything. The button `Build All` from within RStudio triggers this process.

![Build Pane in RStudio](Images/build-pane.png)

@@

However, succesful reproducibility relies on the crucial assumption that your computational environments are identically or sufficiently compatible, e.g., all software dependencies are installed (R and all additional R Packages) and no updates or other changes to the computational environment break or alter your analysis. As we will shortly see, ensuring full computational reproducibility requires one further level of documentation, that is, documentation of the computational environment.

@@

Docker is a tool that allows to encapsulate, share, and re-create a computational environment on most operating systems (Windows, macOS & Linux). Docker achieves these goals by setting up a virtual computer, on which it can execute well-defined series of commands (e.g., installing software). It then saves the resulting state of the virtual computer in what is called an "image." This image can be started and execute commands on the virtual computer, e.g., running `Rscript` or `make`. A running instance of an image is called a container. An image can be transferred and executed on any machine that has docker installed. Irrespective of the machine that is executing the container, the computational environment is the same for the programs running inside the container. The most important advantage over traditional virtual machines is that containers are lightweight, that is, they start fast and don't need much storage space. Docker achieves this by reusing large parts of the hosts operating system (how much is varying widely between Linux, macOS & Windows).

With the following example, we demonstrate why it is important to store the computational environment. Generally, with containers we would like to safe-guard against changes to the computational environment with unexpected consequences, e.g., changes in the functionality or default options in packages or even in the R environment itself. While the R programming language is considered stable and much effort is put into backward compatibility, sometimes even basic functions like `sample()` (to randomly sample from a set) change their behavior from one version to another. To ensure reproducibility in analyses based on a computers pseudo random number generator (PRNG), it is good practice to rely on fixed PRNG seeds, which are numeric values that set the PRNG into a deterministic state, i.e., the sequence of random numbers reproduces exactly. Consider the following R-code to randomly draw five numbers between 1 and 10:

```{r,eval=FALSE}
set.seed(1234)
sample(1:10, 5)
```
The usual expectation is that this code delivers the same pseudo-random five numbers regardless of the operating system or R-Version (due to `set.seed()`). Using Docker, we can start an image which contains the R-Version 3.5.0, and execute the code there.

@@

Surprisingly in an image with a more recent version of R (3.6.0), this results in another sample despite the identical seed:

@@

Note that this is the intended behavior and the result of a [bugfix](https://bugs.r-project.org/bugzilla/show_bug.cgi?id=17494) in the random number generator implemented from R 3.6.0 upwards. Now, such changes may render analyses run on previous R versions strictly not reproducible if they contain, e.g., multiple imputations, bootstrapping, simulations studies, graphics with random jitter, Bayesian estimations using sampling algorithms (such as Markov Chain Monte Carlo), or similar techniques that involve random sampling. We would like to illustrate this with a more concrete example (the full R code to reproduce this non-reproducibility is provided in the GitHub repository of this manuscript). Using a simulated dataset with two variables `x` and `y`, we ran a linear regression model with R's `lm()` function regressing `x` on `y`. Using the `boot` package [@boot], we bootstrapped the 95% confidence intervals around the regression coefficient estimate with 1000 bootstrap samples to evaluate whether zero was inside the confidence interval. To make the analysis reproducible, we set a random seed. We ran this once in R 3.5.0:

@@

Then, we ran the identical script with identical seed in R 3.6.1:

@@

As we see from these R outputs, the confidence intervals once include zero and once do not include zero. Please note that one could discuss deeper issues about null hypothesis significance testing here but with this example, we would simply like to stress the point that computational reproducibility in the strict sense requires capturing the full computational environment.

Only rarely does an analysis depend only on base R. Often, a considerable number of packages is required, and the addition of each package will increase the likelihood of breaking reproducibiity with every update to any of the packages of base R. The whole endeavor of reproducibility is therefore at stake, every time an update is rolled out. To ensure long-term reproducibility, Docker replicates the original computational environment of an analysis exactly. Note that we do not intend to advocate that software should not be updated; updates typically promote bugfixes and provide new functionality; but full computational reproducibility is only achieved if the originally used software versions are documented. Among other things, this allows to trace back recent updates as to what change in what package caused a non-reproduction of the original results. Quite to the contrary, with solutions like Docker, it gets more comfortable than ever to safely update to new versions, by just changing the R Version number of the Docker image. This convenience is possible because of the efforts of the [rocker project](https://GitHub.com/rocker-org/rocker) [@boettigerIntroductionRockerDocker2017a], which provides Docker images preconfigured with an R-installation of given R versions. These packages are taken from MRAN [@revolutionanalyticsReproducibilityUsingFixed2019], a repository for R packages fixed to the last date the R-version of the image was the most recent. Building upon these rocker-images, researchers can easily build their own Docker images with all required R packages. The rocker project also provides images that include RStudio (`rocker/rstudio`), the `tidyverse` package (`rocker/tidyverse`) and the R Markdown package with LaTeX (`rocker/verse`). Because our workflow relies on R Markdown, we suggest using the `rocker/verse` image (which also contains `rstudio` & `tidyverse`). These images are stored on [dockerhub](https://hub.docker.com) [@dockerinc.DockerHub2019].

@@

The `FROM` statement specifies which docker image to use, in this case, the `rocker/verse` image with the tag 3.6.1 (referring to the R version 3.6.1). The `RUN` statement describes a command to execute, in this case, to run an R script `install2.r` which is available on all rocker images, to install the specified packages (here, `pacman`,  `here` & `pander`). A Dockerfile allows more than one `RUN` statement, executing arbitrary system commands. Those `RUN` statements can install dependencies that are not an R package, e.g., other programming languages like python or Matlab. The `WORKDIR` statement is not strictly necessary but saves time spent writing the working directory. The command `docker build -t image-name` creates an image named `image-name` from the Dockerfile in the project. A way to automatically find out the dependencies and genereate a docker image from them is provided in the [`liftR`](https://cran.r-project.org/web/packages/liftr/vignettes/liftr-intro.html) package [@liftr].

There are two principled ways to share a docker image, either by sharing the Dockerfile that creates the image, or by sharing the image itself, e.g., through a service like [dockerhub](https://hub.docker.com). While both ways guarantee a replicable computational environment, sharing the Dockerfile is more transparent; in our workflow, we can use Git to track changes in the Dockerfile (such as updates to dependencies).

To start a container and run commands in a container. Both are based on the `docker run` command. The first way is to run a command inside the container. The call takes the form:

@@

The `--rm` flag means to not save the state of the docker after the command finishes in an container. The `-it` flag tells Docker to accept inputs and return outputs to the terminal. Interacting with docker this way is especially useful when using `R` as a command because then you have access to the standard R terminal. This is the command to start an interactive R session inside a Docker image called `reproducible-research`:

@@

![R-terminal running inside docker](Images/docker-r-terminal.png)

The second way is to start the container in the background (not supplying a specific command) and to interact with the container via the webbrowser and the rstudio server instance that runs in it. In order to do so, you need to supply a password to log into Rstudio server (`-e PASSWORD=<YOUR_PASS>`) and open a port (`-p 8787:8787`).

@@

The adress to connect to the Rstudio server is your ip adress (or, on Linux also `localhost`) in this scheme: `<IPADRESS>:8787`. This offers a fully functioning Rstudio instance that runs in the image but is accessible through a local web browser.

By default, programs inside the container cannot access files on the local computer; to enable access, a local folder needs to be linked explicitly:

@@

Since Docker commands tend to grow long and become tedious to type manually, we advocate to use some automatic way to generate them. Fortunately, one can use `make` to automatically generate the `docker` commands, e.g. the (simplified) Makefile for this paper allows command after `$(run)` to be conditionally passed through Docker if one calls `make DOCKER=TRUE` (otherwise, they are run locally):

```{bash, eval=FALSE}
projekt := reproducible-research
current_dir := $(shell git rev-parse --show-toplevel)
home_dir := $(shell git rev-parse --show-toplevel)
uid = $(shell id -u)

@@

Other than on R, RStudio, and R Markdown, our workflow relies on three pieces of software from outside the R environment: `git`, `make`, and `docker`. The installation process of these software packages depends on the operating system, e.g., on macOS, Make is allways availible, or on Linux, Git and Make are often allready installed. In the following section, we share what have experienced as the easiest way to install those packages. However, installation processes may be subject to change and we advise readers to also consult the documentations of the packages.

@@

Windows systems typically require the biggest efforts to install all necessary software. Note that you have to have Windows Pro, Enterprise, Education or Server installed, as Microsoft prevents the use of Docker on Windows Home. There is a package manager for Windows called chocolately, which you can install from: [https://chocolatey.org/install](https://chocolatey.org/install), providing all dependencies.
After having installed chocolately (and restarted the computer), all dependencies can be installed in an **admin terminal** (Windows-Key, then type `cmd`, right-click *Run as administrator*) via:

@@

To use `docker` you need to start Docker Desktop.

@@

There is a host of diffferent Linux distributions and almost as many package managers. Still, to our knowledge there is no (recent) Linux edition, where `git`, `make` and `docker` are not availible. For example, in Ubuntu Linux, installation is straightforward using the shipped package manager:

>>>>

---
title             : "A Reproducible Data Analysis Workflow with R Markdown, Git, Make, and Docker"
shorttitle        : "Reproducible Data Analysis Workflow"

@@

affiliation:
  - id            : "1"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "2"
    institution   : "Humboldt-Universität zu Berlin"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany & London, U.K." 
authornote: |
   We are grateful to Julia Delius for her helpful assistance in language and style editing. This paper is fully reproducible using the workflow described in this paper. All materials can be found on the accompanying GitHub repository: [https://github.com/aaronpeikert/reproducible-research/](https://github.com/aaronpeikert/reproducible-research/). To provide feedback or any other type of comment or questions regarding our workflow, please add an issue to the GitHub repository of our paper at [https://github.com/aaronpeikert/reproducible-research/issues](https://github.com/aaronpeikert/reproducible-research/issues).

@@

In this tutorial, we describe a workflow to ensure long-term and cross-platform reproducibility of
R-based data analyses. To us Reproducibility means that statistical results as well as their generating computational workflows are reported fully, and transparently, and remain sustainably available, such that a given workflow can be re-run by a different person later in time and the results will be identical to the ones intially reported [@herouxCompatibleReproducibilityTaxonomy2018; @claerboutElectronicDocumentsGive1992]. The need to ensure reproducibility directly follows from commonly accepted rules of good scientific practice [such as the guidelines of the German Research Foundation, @dfg2019]. Ensuring reproducibility is a prerequisite for replicability and a means to increase trustworthiness of empirical results [@epskamp2019rep].

Here, we combine four software tools, whose interplay can guarantee full computational reproducibility of data analyses and their reporting. There are various ideas on how to enhance reproducibility [@piccoloToolsTechniquesComputational2016], four of which we believe to be particularly important: literate programming: [@ruleTenSimpleRules2019], version control [@barbaHardRoadReproducibility2016], dependency management [@askrenUsingMakeReproducible2016], and containerization [@clyburne-sherinComputationalReproducibilityContainers2018]. We argue that only a workflow using all four concepts in unison can guarantee confidence in reproducing a scientific report. Various implementations of these concepts exis6, but we consider the following four best suited for the R environment [@r2019]: for literate programming, R Markdown [@rmarkdown2019], for version control, Git [@chaconProGit2014], for dependency management, Make [@feldmanMakeProgramMaintaining1979], and for containerization, Docker [@merkelDockerLightweightLinux2014]. Each of these software solutions serves a valuable meta-scientific goal (reproducibility) and increases the researches' productivity. They all are highly flexible and powerful, so that mastering them fully, requires a considerable amount of practice. However, for our purposes, it is sufficient to master a valuable minimal subset of functions to ensure reproducibility of scientific analyses.

@@

This paper is fully reproducible using the workflow described in this paper. All materials, including the manuscript source and all R codes and files used to generate this manuscript, can be found on the accompanying GitHub repository: [https://github.com/aaronpeikert/reproducible-research/](https://github.com/aaronpeikert/reproducible-research/).

@@

The first step towards reproducibility is to create an R script or R Markdown file as primary entry point for the analysis that runs on a local computer without error and performs the main statistical analyses. Next, one needs to make sure that all files relevant for the analysis can be moved to another computer. To this end, it is recommendable that all files are within one folder (or enclosed subfolders within it) and all paths are relative to that folder. A robust solution to the problem of making sure that file access does not break across computing platforms are [Rstudio projects](https://r4ds.had.co.nz/workflow-projects.html) and the `here` package [@here2017].

@@

The folder where all the files reside that you need for analysis (code and data), is referred to as a 'project' (or sometimes as a 'research compendium'). Working with projects is particularly convinient with RStudio, an integrated development environment (IDE) for R. It is useful to organize a data analysis project in a way that strictly segregates (raw) data and code by placing them in directories called `data` and `R` [see Section 4 in @marwickPackagingDataAnalytical2018]; there are also tools that automatize the standardized creation of folder structures such as `workflowr` [@workflowr].

Sometimes external requirements make it impossible fot the data to be stored and shared together with the scripts. In most of the cases we have seen, these are either space constraints or privacy considerations. In these cases, of course, unrestricted reproducibility is not guaranteed. If dividing data and scripts is unavoidable, we recommend validating all data files using checksums (also called a "hash", e.g., using the functions provided in package `digest` [@digest2019]) before analyzing them. To do this, a checksum must be created and stored at the time of the original analysis. At the time of reproduction, the current checksum must then be compared with the stored checksum to ensure data integrity.

@@

The translation of computational results into a human-readable summary, for example into a technical report, a presentation, or a journal manuscript, is time-consuming and error-prone. Typical errors result from copy\&paste mistakes, erroneous rounding, or missed updates of the manuscript when the computer code has changed. In order to not only create fully reproducible results but also fully reproducible reports, one needs to resort to the Literate Programming paradigm [@knuth1984literate], in which human-readable language and computer code are intermingled to create dynamic documents whose order follows the logic of thought rather then the order of the computer. R Markdown is a simple markup language to create dynamic documents with embedded chunks of R code that can be exported to standard formats such as documents (docx, rtf, epub), presentations (ppt, html) or websites (html) using the `knitr` package [@xie2015; @xie2019]. Several packages extend the functionality of knitr. Of particular note are the `papaja` package [@papaja2018], which offers additional functions to enable American Psychological Association (APA) style document formatting, including a journal-style final typeset format, and the `stargazer` package [@stargazer2018], which provides journal-ready tables and reports of statistical models.  Figure 1 illustrates R Markdown syntax using the `papaja` package and Figure 2 shows the resulting rendered document.

@@

Fundamentally, reproducibility means that results remain identical if neither the script nor th data have changed. It is often not trivial to find out whether anything in a project has changed over time and if so, to "go back in time." The Git program enables you to do both. A good mental model for Git is that it takes a sequence of snapshots of all files it is supposed to track. In the language of Git, these snapshots are "commits." A commit represents a complete copy of the state of all tracked files. Each commit has a unique identifier (a hash) and a human-readable description (commit message). Going back to one state is as easy as traversing the history of all commits and switching to a given one; it is possible to visually track changes between different versions. Finding a specific version can be made even easier by tagging snapshots, e.g., as "submission," "preprint," or "published." This collection of snapshots is called a "repository," which ideally tracks your entire R project.

@@

To keep track of all changes on your local computer, you only need to use `git add` and `git commit` or `git commit -a` to add and commit at the same time. These commands need to be executed in the terminal, which you can access from within Rstudio (`Shift + Alt + R`).  Rstudio also offers a graphical user interface for Git and, quite likely, this interface is sufficient for almost every interaction with Git.

@@

Git also makes it particularly easy to share and collaborate on a project with other researchers. A popular service for sharing materials via Git is [GitHub](https://github.com). Just sharing Git repositories with the public is always free, private repositories (only visible to persons you invite) are [free for researchers](https://help.github.com/en/articles/applying-for-an-educator-or-researcher-discount). After creating a user account, one can create a new repository and GitHub provides information on how to upload your repository from the terminal, e.g., for our repository (here with user name 'aaronpeikert' and repository name 'reproducible-research'):

@@

`git push` or the green upward arrow in the Git pane (see Figure 3) uploads local updates. To download this Git repository to another computer, type into the terminal:

@@

Git and GitHub can do even more to support you when collaborating with fellow researchers, for example, by providing a web interface to track issues and their status (open/closed/resolved) and further means to manage and merge multiple, parallel versions of code (such as branches, pull requests, or merges), but this is beyond the scope of this tutorial. In particular, GitHub's issue management can be leveraged as a post-publication platform to discuss manuscripts and their results [^1].
Another benefit of using Git and GitHub is that experimentation is highly encouraged since you can go back to any state quickly and even when you lose access to the file on your computer, everything is backed up on the GitHub servers. Further, one can reduce the liklihood of dead code accumulating (e.g., lines that have been commented out) because it is safe to simply remove unneeded code blocks and track the removal in git.

@@

Even when you have obtained a given version of a project with the aim to reproduce reported result and you can confirm that this version is unchanged, you may not know exactly how to reproduce the results because it may be unclear which files should be executed in which order. This is particularly the case when complex preprocessing pipelines are part of the computation and/or external dependencies are many. Handling such dependencies is easy with Make because it allows you to create (computational) recipes to (re-)create files.

@@

The first line can be interpreted as follows: To create `analysis.pdf`, the file `analysis.Rmd`, which depends on `data/clean.csv`, needs to be rendered. This dependency is itself a target. To create `data/clean.csv`, `R/clean.R` and `data/iris.csv` are needed. If you type `make analysis.pdf`, `make` first checks whether the dependencies do exist and, if not, creates them. Here, if `data/clean.csv` does not exist, `make` creates it. Also, if one of the dependencies of a target is newer then the target itself, then Make updates everything that directly or indirectly depends on the target. Here, if `data\iris.csv` is newer then `data\clean.csv`, Make will attempt to recreate `data\clean.csv` first and then `analysis.pdf`. If there is a dependency which is missing, and there is no recipe to make it, `make` stops with an error message. It is a convention to have the first target named `all`, which depends on everything. Then the command `make` without any argument automatically creates everything. The button `Build All` from within RStudio triggers this process (see Figure 4).

![Build pane in RStudio](Images/build-pane.png)

@@

However, successful reproducibility relies on the crucial assumption that your computational environments are identically or sufficiently compatible, e.g., all software dependencies are installed (R and all additional R Packages) and no updates or other changes to the computational environment break or alter your analysis. As we will shortly see, ensuring full computational reproducibility requires one further level of documentation, that is, documentation of the computational environment.

@@

Docker is a tool that allows encapsulation, sharing, and re-creation of a computational environment on most operating systems (Windows, macOS & Linux). Docker achieves these goals by setting up a virtual computer, on which it can execute well-defined series of commands (e.g., installing software). It then saves the resulting state of the virtual computer in what is called an "image." This image can be started and execute commands on the virtual computer, e.g., running `Rscript` or `make`. A running instance of an image is called a container. An image can be transferred and executed on any machine that has docker installed. Irrespective of the machine that is executing the container, the computational environment is the same for the programs running inside the container. The most important advantage over traditional virtual machines is that containers are lightweight, that is, they start rapidly and do not need much storage space. Docker achieves this by reusing large parts of the host's operating system (the extend varies between Linux, macOS and Windows).

With the following example, we demonstrate why it is important to store the computational environment. Generally, with containers we would like to safeguard against changes to the computational environment with unexpected consequences, e.g., changes in the functionality or default options in packages or even in the R environment itself. While the R programming language is considered stable and much effort is put into backward compatibility, even basic functions like `sample()` (to randomly sample from a set) sometimes change their behavior from one version to another. To ensure reproducibility in analyses based on a computer's pseudo random number generator (PRNG), it is good practice to rely on fixed PRNG seeds, which are numeric values that set the PRNG into a deterministic state, i.e., the sequence of random numbers reproduces exactly. Consider the following R code to randomly draw five numbers between 1 and 10:

```{r,eval=FALSE}
set.seed(1234)
sample(1:10, 5)
```

The usual expectation is that this code delivers the same pseudo-random five numbers regardless of the operating system or R Version (due to `set.seed()`). Using Docker, we can start an image which contains the R Version 3.5.0, and execute the code there.

@@

Surprisingly in an image with a more recent version of R (3.6.1), this results in another sample despite the identical seed:

@@

Note that this is intended behavior as it is the result of a [bugfix](https://bugs.r-project.org/bugzilla/show_bug.cgi?id=17494) in the random number generator implemented from R 3.6.0 upwards. Now, such changes may strictly render analyses run on previous R versions not reproducible if they contain, e.g., multiple imputations, bootstrapping, simulations studies, graphics with random jitter, Bayesian estimations using sampling algorithms (such as Markov Chain Monte Carlo), or similar techniques that involve random sampling. We would like to illustrate this with a more concrete example (the full R code to reproduce this non-reproducibility is provided in the GitHub repository of this manuscript). We ran a linear regression model on a simulated dataset with two variables `x` and `y` with R's `lm()` function regressing `x` on `y`. Using the `boot` package [@boot], we bootstrapped the 95% confidence intervals around the regression coefficient estimate with 1000 bootstrap samples to evaluate whether zero was inside the confidence interval. To make the analysis reproducible, we set a random seed. We ran this once in R 3.5.0:

@@

Then, we ran the identical script with the identical seed in R 3.6.1:

@@

As we see from these R outputs, the confidence intervals once include zero and once do not. Please note that one could discuss deeper issues about null hypothesis significance testing here but with this example, we would simply like to stress the point that computational reproducibility in the strict sense requires capturing the full computational environment.

Only rarely does an analysis depend on base R only. Often, a considerable number of packages is required, and the addition of each package will increase the likelihood of breaking reproducibility with every update to any of the packages of base R. The whole endeavor of reproducibility is therefore at stake every time an update is rolled out. To ensure long-term reproducibility, Docker replicates the original computational environment of an analysis exactly. Note that we do not intend to advocate that software should not be updated; updates typically promote bugfixes and provide new functionality; our point is that full computational reproducibility is only achieved if the originally used software versions are documented. Among other things, this makes it possible to trace back recent updates as to what change in what package caused a non-reproduction of the original results. Quite to the contrary, with solutions like Docker, it gets easier than ever to safely update to new versions just by changing the R version number of the Docker image. This convenience is possible because of the efforts of the [`rocker` project](https://GitHub.com/rocker-org/rocker) [@boettigerIntroductionRockerDocker2017a], which provides Docker images preconfigured with an R-installation of given R versions. These packages are taken from MRAN [@revolutionanalyticsReproducibilityUsingFixed2019], a repository for R packages fixed to the last date on which the R version of the image was the most recent. Building upon these rocker images, researchers can easily build their own Docker images with all required R packages. The rocker project also provides images that include RStudio (`rocker/rstudio`), the `tidyverse` package (`rocker/tidyverse`) and the R Markdown package with LaTeX (`rocker/verse`). Because our workflow relies on R Markdown, we suggest using the `rocker/verse` image (which also contains `rstudio` & `tidyverse`). These images are stored on [dockerhub](https://hub.docker.com) [@dockerinc.DockerHub2019].

@@

The `FROM` statement specifies which docker image to use, in this case, the `rocker/verse` image with the tag 3.6.1 (referring to the R version 3.6.1). The `RUN` statement describes a command to execute, in this case, to run an R script `install2.r` which is available on all rocker images, to install the specified packages (here, `pacman`,  `here` & `pander`). A Dockerfile allows more than one `RUN` statement, executing arbitrary system commands. Those `RUN` statements can install dependencies that are not an R package, e.g., other programming languages like python or Matlab. The `WORKDIR` statement is not strictly necessary but saves time spent on writing the working directory. The command `docker build -t image-name` creates an image named `image-name` from the Dockerfile in the project. A way to identify the dependencies automatically and generate a docker image out of them is provided in the [`liftR`](https://cran.r-project.org/web/packages/liftr/vignettes/liftr-intro.html) package [@liftr].

There are two principled ways to share a docker image; either by sharing the Dockerfile that creates the image, or by sharing the image itself, e.g., through a service like [dockerhub](https://hub.docker.com). While both ways guarantee a replicable computational environment, sharing the Dockerfile is more transparent; in our workflow, we can use Git to track changes in the Dockerfile (such as updates to dependencies).

There are two options to interact with a container. Both options are based on the `docker run` command. The first way is to run a command inside the container. The call takes the form:

@@

The `--rm` flag means that the state of the docker after the command finishes in an container is not saved. The `-it` flag tells Docker to accept inputs and return outputs to the terminal. Interacting with docker this way is especially useful when using `R` as a command because you then have access to the standard R terminal. This is the command to start an interactive R session inside a Docker image called `reproducible-research`:

@@

![R terminal running inside docker](Images/docker-r-terminal.png)

The second option is to start the container in the background (not supplying a specific command) and to interact with the container via the webrowser and the RStudio server instance running in it. In order to do so, you need to supply a password to log into the Rstudio server (`-e PASSWORD=<YOUR_PASS>`) and open a port (`-p 8787:8787`).

@@

The adress to connect to the RStudio server is your IP address (or `localhost` Linux) in this scheme: `<IPADRESS>:8787`. This offers a fully functioning Rstudio instance that runs in the image but is accessible through a local web browser.

By default, programs inside the container cannot access files on the local computer, thus requiring an explicit link to a local folder to enable access (and on Mac and Windows this also have to be allowed in the settings) :

@@

Since Docker commands tend to grow long and become tedious to type manually, we recomend using some automatic way to generate them. Fortunately, one can use `make` to automatically generate the `docker` commands, e.g. the (simplified) `Makefile` for this paper allows command after `$(run)` to be conditionally passed through Docker if one types `make DOCKER=TRUE` (otherwise, they are run locally):

```{bash, eval=FALSE}
projekt := $(notdir $(CURDIR))
current_dir := $(CURDIR)
home_dir := $(current_dir)
uid = --user $(shell id -u)

@@

Other than on R, RStudio, and R Markdown, our workflow relies on three pieces of software from outside the R environment: `git`, `make`, and `docker`. The installation process of these software packages depends on the operating system, e.g., on macOS, Make is always available, while on Linux, Git and Make are often already installed. In the following section, we share what what we consider the easiest way to install those packages. However, installation processes may be subject to change, and we advise that readers also consult the documentations of the packages.

@@

Windows systems typically require the biggest efforts to install all necessary software. Note that you have to have Windows Pro, Enterprise, Education or Server installed, as Microsoft prevents the use of Docker on Windows Home. There is a package manager for Windows called chocolately, which you can install from: [https://chocolatey.org/install](https://chocolatey.org/install), providing all dependencies.
Having installed chocolately (and restarted the computer), all dependencies can be installed in an **admin terminal** (Windows key, then type `cmd`, right-click *Run as administrator*) via:

@@

To use `docker` you need to start Docker Desktop. In the settings of docker you have to allow the sharing of your drive. Docker on Windows require the path to be special; therefore you need to hand edit the `Makefile` and set `current_path` to the current directory and use `make all DOCKER=TRUE WINDOWS=TRUE`. We hope that future releases of Docker for Windows will not require that workaround.

@@

There is a host of different Linux distributions and almost as many package managers. Still, to our knowledge, there is no (recent) Linux edition, that does not include `git`, `make` and `docker`. For example, in Ubuntu Linux, installation is straightforward using the shipped package manager:
